---
title: "Writing a great story for data science projects "
subtitle: "This is a Report Template Quarto"
author: "Aliyah Hunter & Sophia Craig (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report! I am adding this to show updates from RStudio.
:::

## Introduction

The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

<!-- -->

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*. Topology can be used in machine 
learning [@adams2021topology]

For Symbolic Regression [@wang2019symbolic] *This is my work and I want
to add more work...*

Cite new paper [@su2012linear]

Classification of Knee OA was studied here [@guida2021knee]. This is another example of LCA [@weller2020latent]


## Literature Review
Other empirical studies have shown that the Cox PH model can be used well beyond its origin in clinical survival analysis. The studies reviewed for this project applied the model to academic performance, childbirth, agricultural economics, website use, and credit risk. Thus, the model could be used in any setting that has time-to-event data with censoring. In a study on medical student attrition, a stepwise Cox model was used to examine which academic and demographic factors predicted when a student would experience academic difficulty. Researchers found that curriculum track, MCAT verbal reasoning scores, and early course failures were the strongest predictors, and that the risk peaked at a specific point in the training timeline [@chau-kuang2010]. In a study on the duration of childbirth, Cox PH was used to examine whether a specific birth-preparation method changed the time it took to deliver. The study found that this method significantly shortened the time to delivery compared to standard practice [@wuryandari2018]. Cox PH has also been used to look at economic behavior. Specifically, [@hidayat2022] modeled how long it took rice farmers to pay an agricultural insurance premium, using the Efron partial-likelihood method and stepwise variable selection. They found that education level, off-farm employment, and production costs affected how quickly farmers acted. Also, [@soobramoney2023] treated website user drop-off as a survival problem, in which “survival” meant viewing at least three pages. Both Cox PH and a survival random forest model were used, and browser type, landing page, and search-referral status were found to be key hazards. This shows that the framework is not limited to health and social science, and it can be applied to behavioral and business data as well. Together, these studies show that all the model needs is a duration variable and a binary event or censoring indicator. Also, the Cox model can handle multiple covariates flexibly. Therefore, it is a tool that could be useful wherever the timing of an outcome is of interest. For these reasons, Cox PH was chosen for the recidivism data in this study, which has this same time-to-event structure.

Another theme in the literature is that researchers should test that a covariate's effect on hazard stays constant over time. [@bellera2010] warned that predictors can look strong early in follow-up and then weaken considerably over time. Using Schoenfeld residuals, they showed that hormone receptor status in breast cancer patients had this kind of time-varying effect. A standard Cox model would have hidden this finding. [@abadi2014] reached a similar conclusion in a separate and larger breast cancer survival study. They found that the proportional hazards assumption held differently across age and cancer-stage subgroups. Therefore, stratified Cox models were used to account for this difference between the groups. Also, [@lee2023] made the same point in a more general methodological review. This review noted that hazard or survival curves that overlap on a graph are a practical sign that the PH assumption has been violated. Failing to detect a violation could lead to misleading conclusions about the true relationship between a covariate and the outcome.

Some of the reviewed studies went further and compared modified Cox specifications that were designed to address PH violations. [@iketle2024] compared the stratified Cox model, which allows a different baseline hazard for each level of a violating covariate, against the extended Cox model, which puts time-varying covariates directly into the hazard function. This was done using Monte Carlo simulations across different censoring rates, sample sizes, and survival distributions. They found that the extended model generally performed better than the stratified model when a true time-varying effect was present. However, they also noted that there were practical limitations in specifying the correct functional form for that time dependence. [@10689732] also emphasized that the Cox model's main value is that it can isolate each covariate's independent contribution to hazard while controlling for the others. However, they cautioned that the PH assumption itself, along with the interpretation of complex covariate interactions, remain persistent limitations of this approach. Also, [@dirick2017] took this comparative approach even further in a credit-risk context. They benchmarked Cox PH against accelerated failure time models, spline-based Cox extensions, and mixture cure models, the last of which are designed to account for a subpopulation that will never experience the event at all. This study found that Cox PH with penalized splines consistently performed better than fully parametric alternatives across ten credit datasets. This reinforced that Cox PH is flexible even when the relationship between covariates and hazard is nonlinear.

Taken together, the literature suggests that a standard Cox PH model be used, then, the proportional hazards assumption will be tested using Schoenfeld residuals. If violations are found they will be addressed through stratification or time-varying covariate specifications.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
