---
title: "Writing a great story for data science projects "
subtitle: "This is a Report Template Quarto"
author: "Aliyah Hunter & Sophia Craig (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)


## Introduction
In many studies, the outcome of interest is not only whether an event occurs, but also how long it takes for that event to happen. For example, researchers may be interested in how long a patient survives after treatment, when a borrower defaults on a loan, or when a user leaves a website. Survival analysis considers both whether an event occurs and the time until it occurs, while accounting for censored observations.

The Cox proportional hazards model is one of the most commonly used methods for analyzing time-to-event data. The model allows researchers to examine how different variables are related to the risk of experiencing an event over time. Because of this, Cox regression can be applied to many different types of research questions. Previous studies have used Cox models to examine the duration of the birth process [@Wuryandari2018], the time until farmers are willing to pay agricultural insurance premiums [@Hidayat2022], website drop-offs [@Soobramoney2023], credit default [@Dirick2017], and survival following cancer treatments [@Abadi2014]. Although these studies focus on very different outcomes, they all use the same general idea of examining the time until a specific event occurs.

An important part of using the Cox model is the proportional hazards assumption. This assumption means that the effect of a predictor on the hazard is expected to remain proportional over time. For example, if one group has a higher risk of experiencing an event than another group, the model assumes that the relative difference between those groups remains consistent throughout the study period. If that relationship changes over time, a single hazard ratio may not accurately describe the effect of the predictor.

Because of this, the proportional hazards assumption should be evaluated before the results of a Cox model are interpreted. However, this does not always happen in practice. Reviews of pediatric leukemia and total joint arthroplasty studies found that the proportional hazards assumption was often not reported or was not adequately evaluated [@Kuitunen2021]; [@Kuitunen2023]. Bellera et al. also emphasized that researchers should investigate whether effects change over time rather than automatically assuming that proportional hazards are present [@Bellera2018].

When the proportional hazards assumption is violated, the standard Cox model may not be the best representation of the data. Researchers can use modifications such as stratification or time-dependent effects to account for relationships that change throughout the study period. However, the performance of these approaches can also depend on factors such as sample size, censoring, and the distribution of survival times [Iketle2024]. Overall, the Cox proportional hazards model provides a useful and flexible way to analyze time-to-event outcomes, but its results depend on appropriately accounting for censored observations and evaluating whether the proportional hazards assumption is reasonable for the data.


### Literature Review
#### Cox PH model across domains
Empirical studies have shown that the Cox PH model can be used well beyond its origin in clinical survival analysis. The studies reviewed for this project applied the model to academic performance, childbirth, agricultural economics, website use, and credit risk. Thus, the model could be used in any setting that has time-to-event data with censoring. In a study on medical student attrition, a stepwise Cox model was used to examine which academic and demographic factors predicted when a student would experience academic difficulty. Researchers found that curriculum track, MCAT verbal reasoning scores, and early course failures were the strongest predictors, and that the risk peaked at a specific point in the training timeline [@chau-kuang2010]. In a study on the duration of childbirth, Cox PH was used to examine whether a specific birth-preparation method changed the time it took to deliver. The study found that this method significantly shortened the time to delivery compared to standard practice [@wuryandari2018]. [@hidayat2022] applied Cox PH to model economic behavior, using the Efron partial-likelihood method and stepwise variable selection to examine how long it took rice farmers to pay an agricultural insurance premium. They found that education level, off-farm employment, and production costs affected how quickly farmers acted.[@soobramoney2023] treated website user drop-off as a survival problem, in which “survival” meant viewing at least three pages. Both Cox PH and a survival random forest model were used, and browser type, landing page, and search-referral status were found to be key hazards. This shows that the framework is not limited to health and social science, and it can be applied to behavioral and business data as well. Together, these studies show that all the model needs is a duration variable and a binary event or censoring indicator. Also, the Cox model can flexiby handle multiple covariates. Therefore, it is a tool that could be useful wherever the timing of an outcome is of interest. For these reasons, Cox PH was chosen for the recidivism data in this study, which has this same time-to-event structure.

#### Proportional hazards assumption
Another theme in the literature is that researchers should test that a covariate's effect on hazard stays constant over time. [@bellera2010] warned that predictors can look strong early in follow-up and then weaken considerably over time. Using Schoenfeld residuals, they showed that hormone receptor status in breast cancer patients had this kind of time-varying effect. A standard Cox model would have hidden this finding. [@abadi2014] reached a similar conclusion in a separate and larger breast cancer survival study. They found that the proportional hazards assumption held differently across age and cancer-stage subgroups and fitting stratified Cox models to account for this difference between the groups. Also, [@lee2023] made the same point in a more general methodological review. This review noted that hazard or survival curves that overlap on a graph are a practical sign that the PH assumption has been violated. Failing to detect a violation could lead to misleading conclusions about the true relationship between a covariate and the outcome. For recidivism data, a covariate's effect on the hazard of reincarceration could be concentrated in the early months, therefore the proportional hazards assumption will be tested.

#### Handling non-proportional hazards
Several reviewed studies compared modified Cox specifications designed to address PH violations. [@iketle2024] compared the stratified Cox model, which allows a different baseline hazard for each level of a violating covariate, against the extended Cox model, which puts time-varying covariates directly into the hazard function. This was done using Monte Carlo simulations across different censoring rates, sample sizes, and survival distributions. They found that the extended model performed better than the stratified model when a true time-varying effect was present. However, they noted that there were practical limitations in specifying the correct functional form for that time dependence. [@msk2024] emphasized that the Cox model's main value was that it could isolate each covariate's independent contribution to hazard while controlling for the others. However, they cautioned that the PH assumption itself, along with the interpretation of complex covariate interactions, remained a limitations of this approach. [@dirick2017] took this comparative approach even further in a credit-risk context. They benchmarked Cox PH against accelerated failure time models, spline-based Cox extensions, and mixture cure models (designed to account for a subpopulation that will never experience the event at all) across ten datasets. This study found that Cox PH with penalized splines consistently outperformed fully parametric alternatives, reinforcing the model's flexibility even when covariate-hazard relationships are nonlinear.

Overall, the Cox PH is well suited to the structure of the Iowa data. This study will begin with a standard Cox PH model, tests the proportional hazards assumption formally using Schoenfeld residuals, and addresses any violations dectected through stratification or time-varying covariate specifications.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

The Cox Proportional Hazards model is a regression method used to analyze the relationships between variables and the time until a specific event occurs. Survival analysis accounts for both the occurrence of an event and the length of time before the event, while accounting for censored observations. 

The Cox model expresses an individual's hazard at time $t$ as

$$
\lambda(t \mid X) = \lambda_0(t) \exp(\beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p)
$$
$$
\text{HR} = \exp(\beta_j)
$$

where $\lambda_0(t)$ is the baseline hazard, $\boldsymbol{\beta}$ is a vector of regression coefficients, and $X$ is a vector of time-fixed covariates [@bellera2010]. Exponentiating a coefficient produces a hazard ratio, which describes the relative risk of experiencing the event while holding other variables constant. A hazard ratio greater than one indicates an increased risk. For example, Cox models have been used to study the duration of the birth process [@wuryandari2018], the time until farmers are willing to pay agricultural insurance premiums [@hidayat2022], website drop-offs [@soobramoney2023], credit default [@dirick2017], and survival following cancer treatments [@abadi2014]. These applications demonstrate how the model can be used when the outcome of interest involves the time until a specified event. The general process for conducting a Cox proportional hazards analysis is 1. Define the event of interest, survival time, and censoring indicator, 2. Select the explanatory variables that may affect the event hazard, 3. Fit the Cox proportional hazards model and estimate its coefficients, 4. Convert the coefficients into hazard ratios and calculate their confidence intervals, 5. Test whether the proportional hazards assumption is satisfied, 6. Evaluate model fit and interpret the estimated effects within the context of the research question [@lee2023].

The central assumption of the Cox model is that the hazard ratio associated with a predictor remains constant throughout the observation period. In other words, the model assumes that the hazards for two individuals or groups are proportional over time. The assumption can be evaluated using Kaplan-Meier curves, log-minus-log plots, Schoenfeld residuals, or interactions between predictors and time. However, reviews of pediatric leukemia and total joint arthroplasty research found that the proportional hazards assumption was frequently unreported or poorly evaluated [@kuitunen2021]; [@kuitunen2023]. Time-varying effects should be investigated as a regular part of Cox regression analysis rather than assuming that proportionality automatically holds [@bellera2010].

One limitation of the Cox proportional hazards model is its assumption that the hazard ratio for each predictor remains constant throughout the study period. If a predictor has a stronger effect at one point in time than another, this assumption is violated. The model may then produce a single hazard ratio that does not accurately represent the changing relationship. Researchers should therefore test the proportional hazards assumption before interpreting the model’s results. When the assumption is violated, modifications such as stratifying the model or including time-dependent effects may be necessary. The performance of these approaches can vary according to the sample size, censoring rate, and distribution of survival times [@iketle2024]. Although the Cox proportional hazards model is flexible, reliable results depend on properly accounting for censored observations and verifying that the proportional hazards assumption is appropriate.

## Analysis and Results

### Data Exploration and Visualization
The Iowa Prison Recidivism Status dataset contains case-level records of individuals released from Iowa prisons through parole, work release, or discharge, beginning with the FY2016 cohort. It covers releases from July 2015 through June 2022 and tracks whether each individual was reincarcerated. Available variables include age, sex, race, the original offense, the convicting offense on recidivism if one occurred, and a risk classification.

This analysis looks at reincarceration outcomes for 31,703 individuals who were followed for up to three years after release. The overall reincarceration rate was 36.9%, but this rate varied across demographic and offense-based subgroups. Violent offenders had the lowest rate at 29.2%, while individuals convicted of "other" offenses had the highest at 50.3%. This goes against a common assumption that more severe offenses lead to worse reentry outcomes. The timing of reincarceration also showed a clear pattern. Most returns happened within the first six months after release, and the rate of return leveled off after that point.

Differences appeared across several other groups as well. Males were reincarcerated at a rate about 8 percentage points higher than females (38.0% vs. 30.0%). American Indian or Alaska Native individuals had the highest rate of any racial group at 44.7%. Risk classification separated outcomes as intended, with high risk individuals reincarcerated at a rate of 42.0% compared to 31.4% for low risk individuals. However, individuals who were never formally assessed had a lower rate than the low risk group. This might indicate some bias and should be investigated further.

The eight release supervision types in the data also produced different outcomes. Individuals discharged at the expiration of their sentence had survival probabilities staying above 70% across the full three year period. Those paroled with immediate discharge and those released to special sentences had survival probabilities falling below 60% within the first year. Individuals granted parole, who made up the largest group at 20,551, fell in between. These differences suggest that release type may be more than a marker of underlying offender characteristics, and that supervision conditions themselves could shape reentry outcomes.

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(survival)
library(survminer)
library(skimr)  
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
data_raw <- read_csv("ipr_dataset.csv", na = c("", "NA"))

# Data Cleaning
empty_cols <- names(data_raw)[colSums(is.na(data_raw)) == nrow(data_raw)]
data_raw <- data_raw %>% select(-all_of(empty_cols))

df <- data_raw %>%
  mutate(
    event = if_else(!is.na(days_to_recid), 1, 0),
    release_date = as.Date(exit_sup_end_date),
    time_days = if_else(event == 1, days_to_recid, 1095)
  )

df <- df %>% mutate(time_days = if_else(time_days == 0, 0.5, time_days))

df <- df %>%
  unite("exit_combined", exit_type, exit_subtype, sep = " - ", remove = FALSE) %>%
  mutate(
    sex           = factor(sex),
    Race          = factor(Race),
    exit_type     = factor(exit_type),
    exit_sup_reason      = factor(exit_sup_reason),
    risk_ranking  = factor(
      if_else(is.na(risk_ranking), "Not assessed", risk_ranking),
      levels = c("Low", "Moderate", "High", "Not assessed")
    ),
    age_c         = age - mean(age, na.rm = TRUE)
  )
df <- df %>% filter(!is.na(sex))
```

```{r}
df %>%
  filter(event == 1) %>%
  ggplot(aes(x = survival_time_months)) +
  geom_histogram(binwidth = 1, fill = "#d1495b", color = "white") +
  labs(title = "When Does Reincarceration Occur?",
       x = "Months to reincarceration", y = "Count") +
  theme_minimal()
```

```{r}
set.seed(20)  

pt <- df %>%
  slice_sample(n = 20) %>%
  arrange(time_days) %>%
  mutate(id = row_number())

plot(NA, xlim = c(0, 1200), ylim = c(0.5, 20.5),
     xlab = "Days since release", ylab = "Individual", yaxt = "n",
     main = "Time to Reincarceration: Events vs. Censoring")
axis(2, at = 1:20, labels = paste0("P", 1:20), cex.axis = 0.7)
abline(v = 1095, lty = 2, col = "grey60")            
text(1095, 20.4, "3-yr cutoff", pos = 2, col = "grey40", cex = 0.8)

for (i in 1:20) {
  segments(0, pt$id[i], pt$time_days[i], pt$id[i], col = "#3a86d4", lwd = 3)
  if (pt$event[i] == 1) {
    points(pt$time_days[i], pt$id[i], pch = 19, col = "#d1495b", cex = 1.4)   
  } else {
    points(pt$time_days[i], pt$id[i], pch = ">", col = "#3a86d4", cex = 1.4) 
  }
}
legend("bottomright", bty = "n",
       legend = c("recidivated (event)", "censored (no return by cutoff)"),
       pch = c(19, 62), col = c("#d1495b", "#3a86d4"))
```
```{r}
km_risk <- survfit(Surv(time_days, event) ~ risk_ranking, data = df)

plot(km_risk, col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"), lwd = 2,
     xlab = "Days since release", ylab = "Probability of remaining in community",
     main = "Time to Reincarceration by Risk Classification")
legend("bottomleft", bty = "n", lwd = 2,
       legend = levels(df$risk_ranking),
       col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"))
```

```{r}
km_risk <- survfit(Surv(time_days, event) ~ exit_type, data = df)

plot(km_risk, col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"), lwd = 2,
     xlab = "Days since release", ylab = "Probability of remaining in community",
     main = "Time to Reincarceration by Offense Type")
legend("bottomleft", bty = "n", lwd = 2,
       legend = levels(df$exit_type),
       col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"))

survdiff(Surv(time_days, event) ~ exit_type, data = df)
```

```{r}
km_risk <- survfit(Surv(time_days, event) ~ sex, data = df)

plot(km_risk, col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"), lwd = 2,
     xlab = "Days since release", ylab = "Probability of remaining in community",
     main = "Time to Reincarceration by Sex")
legend("bottomleft", bty = "n", lwd = 2,
       legend = levels(df$sex),
       col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"))

survdiff(Surv(time_days, event) ~ sex, data = df)
```


```{r}
km_risk <- survfit(Surv(time_days, event) ~ exit_sup_reason, data = df)

plot(km_risk, col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"), lwd = 2,
     xlab = "Days since release", ylab = "Probability of remaining in community",
     main = "Time to Reincarceration by Release Type")
legend("bottomleft", bty = "n", lwd = 2,
       legend = levels(df$exit_sup_reason),
       col = c("#2a9d8f", "#e9c46a", "#d1495b", "grey60"))

survdiff(Surv(time_days, event) ~ exit_sup_reason, data = df)
```


### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
